LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-12-10 article/main.tex
small rewording
remove redundant sentence
replace coling style with icaart style

\documentclass[11pt]{article}
\pdfoutput=1

\usepackage[review]{coling}

@@

\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{balance}
\usepackage{booktabs}
\usepackage[justification=centering]{caption}
\usepackage{cleveref}
\usepackage{comment}
\usepackage{enumitem}
\usepackage{epsfig}
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{inconsolata}
\usepackage[utf8]{inputenc}
\usepackage{interval}
\usepackage{microtype}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subfigure}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{url}
\usepackage{verbatim}
\usepackage{xcolor}
\usepackage{xstring}

@@

\title{LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment}
\authoranon{
\author{Andrei Olar\\
Universitatea Babeș-Bolyai, Cluj-Napoca, România\\
\texttt{andrei.olar@ubbcluj.ro}\\}
}

\begin{document}
\maketitle
\begin{abstract}
    Can we control the writing style of large language models (LLMs) by specifying
    desired linguistic features?
    We address this question by investigating the impact of handcrafted linguistic
    feature (HLF) instructions on LLM-generated text.
    Our experiment evaluates various state-of-the-art LLMs using prompts
    incorporating HLF statistics derived from corpora of CNN articles and Yelp
    reviews.
    We find that LLMs demonstrate sensitivity to these instructions, particularly
    when tasked with conforming to concrete features like word count.
    However, compliance with abstract features, such as lexical variation, proves
    more challenging, often resulting in negative impacts on compliance.
    Our findings highlight the potential and limitations of utilizing HLFs for
    guiding LLM text generation and underscore the need for further research into
    optimizing prompt design and feature selection.
\end{abstract}

@@

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
Examples of HLFs range from simple constructs such as counts or averages of
words, sentences or specific parts of speech~(adjectives, verbs, a.s.o) to more
complex statistics based on heuristics.
All HLFs share the characteristics of computational ease and idempotence.
This makes HLFs very attractive because of their dual potential.
On the one hand, an LLM might be instructed using a prompt containing the value
of an HLF.\@
An example prompt instruction which asks the LLM to conform to a maximum
\textit{sentence count} of 10 would read ``write no more than 10 sentences''.
On the other hand, HLFs can be computed on the text output by an LLM, too.
This dual nature of HLFs allows inspecting how closely does LLM generated text
conform to the goal outlined in the input prompt HLF instruction.\@

@@

The experiment starts with choosing experiment parameter values.
The following parameters have stable values throughout the experiment process: a
selection of HLFs, a corpus containing texts, a target LLM, examples from the
chosen corpus, one input text from the chosen corpus and one input text from
outside the chosen corpus.

Once the parameters have been determined, the HLF statistics are computed on the
text corpus.
Remembering that HLFs are just floating point numbers, the following statistics
are computed for each HLF:~\textbf{min}~(the minimum value), \textbf{max}~(the
maximum value) and \textbf{avg}~(the mean).

The subsequent step involves formulating prompt instructions that incorporate
HLFs.
For each HLF, an instruction is constructed in natural language, adhering to the
previously calculated minimum, maximum, and mean values. For instance, an
instruction corresponding to the total word count HLF might be expressed as
follows:

@@

The prompt templates correspond to each of the two experiment scenarios:
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The static instruction in `\texttt{prompt\_1}' to ``\textit{Write text that
conveys the meaning of the first user prompt}'' ensures that input text meaning
is preserved.
In addition to the primary directive from `\texttt{prompt\_1}',
`\texttt{prompt\_2}' contains instructions concerning the interpretation of the
provided examples, as well as logic to incorporate them.
Both prompt templates contain logic to distinguish between the situation when
HLF instructions were given versus not.

With the preliminary steps completed, baseline HLFs for the target LLM are
computed.
For the first scenario, the LLM is instructed to reword the input text 10 times
by using `\texttt{prompt\_1}' without HLF instructions.
The HLFs in our selection are computed for each generated text, resulting in
a 10-dimensional vector for each HLF.\@
These vectors represent the baseline HLFs for the LLM in the first scenario.

The second scenario baseline values are computed similarly using the
`\texttt{prompt\_2}' template and the examples provided as one of the experiment
parameters.
Remembering that the second scenario has two variants, we obtain two sets of
baseline vectors: a set for input from the corpus and one for input outside the
corpus.
The input from outside the corpus is the same as the input for the first
scenario.

From here on we refer to the HLFs computed using the above process as
\textit{baselines}.
Note that we have 3 sets of baselines: 1 for the first scenario and 2 for the
second scenario.

For each type of baseline, we run 10 text generations using the corresponding
prompt augmented with the HLF instructions computed earlier in the process.
Doing this results in 3 sets of 10-dimensional vectors, each corresponding to
the baselines.
We refer to these vectors as \textit{HLF results}.

The last step is to compare baselines with their corresponding HLF results for
each HLF in our selection, for each scenario and variant.

The comparison result for the first scenario indicates whether LLMs acknowledge
HLF instructions at all.
If they do, the HLF results should be significantly different from their
baseline.

Both variants of the second scenario quantify how similar the HLFs of the texts
generated by the LLM are to the corpus average \textit{because of} the HLF
instructions.
This hypothesis emerges from the assumption that LLMs are able to mimic examples
when they are provided in the system prompt.
In such a case, the presence of HLF instructions would remain among the few
possible explanations for deviations from the baseline.

@@

The LLMs selected for the experiment must represent the state of the art.
The Chatbot Arena~\cite{chiang2024chatbot} leaderboard is essential in our
selection based on this criterion.
To cover more ground we find top performers from different vendors interesting.
The final selection is available in Table~\ref{table-llm}.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
In this experiment we use LFTK~\cite{lftk-2023}, a framework built on top of
spaCy~\cite{spacy} that provides implementations for computing many HLFs.
LFTK categorizes HLFs by domain and family.
We select HLFs from the catalog which is implemented by LFTK based on their
domain and family, with the interest to make a diverse selection with respect to
these attributes.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

@@

An input text from within each corpus is chosen so that it exhibits the closest
HLFs to the averages computed on the corpus.
The choice is made when the text corpora are selected and is stable throughout
all experimental evaluations.

@@

HLFs are real numbers that we can derive from text idempotently.
Plot figures show the HLF on the vertical axis and the trial number on the horizontal
axis.
We plot the baseline using a continuous grey line and the HLF result using a
continuous grenat line.
The minimum and maximum HLF on the corpus are plotted using dotted black lines.
A continuous black line designates the average HLF on the corpus.
We refer to the corpus average as the \textit{target} because the LLM is
instructed to generate text that primarily conforms to this value.

@@

There are exceptions to the general trend observed for the first scenario.
One such exception is highlighted in Figure~\ref{fig-p2-gpt-twords}.
In~\cref{fig-p2-command-r-simpttr,fig-p2-gemini-a-kup-pw} we even observe
regressions in performance with significant differences between baseline and
result.
Even so, these are exceptions and the results obtained in our first scenario are
largely confirmed.

@@

The results shown in Table~\ref{table-prompt-2-ifd-cnn-dailymail} contradict
these expectations.
The most surprising behaviour is exhibited for HLFs which are derived into more
concrete generation instructions, such as the total number of
words~(Figure~\ref{fig-p2-ifd-gemini-twords}).
Additionally, we notice regressive behaviour for some models compared to the
previously explored results in Figure~\ref{fig-p2-ifd-claude3-nuadj}.
Finally, notice that the abstractness of the text generation instructions for
certain HLFs leads to non-compliance in Figure~\ref{fig-p2-ifd-llama3-a-kup-pw}.

@@

The overall sentiment, taking into account the additional results obtained on
the Yelp dataset, is that LLMs do \textit{not} take advantage of the examples
from text corpora for the tasks performed in this experiment.

@@

In terms of our first question --- whether HLF instructions impact an LLM's
writing, the response is affirmative.
Not all LLMs are equally susceptible to HLF instructions, though.
The experiment design involves choices that rely on the presence of a corpus and
the usage of a fixed external input text.
This is relevant as evidentiated by the difference in HLF results obtained using
the Yelp reviews corpus when compared to the HLF results obtained on the CNN
article corpus.
While similarly significant, the HLF results on Yelp reviews are mostly worse
than their corresponding baselines failing to confirm the positive impact of HLF
instructions.
The importance of the choice of the external text is highlighted by the better
HLF results obtained using an input text from the corpus.

There is more doubt in response to how much HLF instructions impact an LLM's
writing.
The metrics we compute dismiss the idea that HLF instructions have a consistent
desired impact on the generated text.
There isn't conclusive evidence that if we improve the LLM's chances of
producing a better baseline, this will result in a smaller difference between
baseline and HLF result relative to the target HLF.
In fact, there is evidence that HLF instructions have an adverse effect when
examining the HLF results obtained on the Yelp reviews dataset.
This is especially true of highly abstract HLF instructions.
LLMs do not seem to follow these instructions well in most scenarios.

>>>>

\documentclass[a4paper,twoside]{article}
\pdfoutput=1

@@

\usepackage{algorithm2e}
\usepackage{array}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsthm}
\usepackage{apalike}
\usepackage{balance}
\usepackage{booktabs}
\usepackage{calc}
\usepackage[justification=centering]{caption}
\usepackage{cleveref}
\usepackage{comment}
\usepackage{enumitem}
\usepackage{epsfig}
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage[bottom]{footmisc}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{inconsolata}
\usepackage[utf8]{inputenc}
\usepackage{interval}
\usepackage{microtype}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{pslatex}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{url}
\usepackage{verbatim}
\usepackage{xcolor}
\usepackage{xstring}

@@

\usepackage{SCITEPRESS}

\begin{document}

\title{LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment}

\author{\authorname{Andrei Olar\sup{1}\orcidAuthor{0009-0006-7913-9276}}
\affiliation{\sup{1}Mathematics and Computer Science Faculty, Babeș-Bolyai University, Cluj-Napoca, România}
\email{andrei.olar@ubbcluj.ro}
}
\keywords{handcrafted linguistic features, large language models, controlled text generation, text style transfer}
\abstract{
    Can we control the writing style of large language models (LLMs) by specifying
    desired linguistic features?
    We address this question by investigating the impact of handcrafted linguistic
    feature (HLF) instructions on LLM-generated text.
    Our experiment evaluates various state-of-the-art LLMs using prompts
    incorporating HLF statistics derived from corpora of CNN articles and Yelp
    reviews.
    We find that LLMs demonstrate sensitivity to these instructions, particularly
    when tasked with conforming to concrete features like word count.
    However, compliance with abstract features, such as lexical variation, proves
    more challenging, often resulting in negative impacts on compliance.
    Our findings highlight the potential and limitations of utilizing HLFs for
    guiding LLM text generation and underscore the need for further research into
    optimizing prompt design and feature selection.
}

\onecolumn \maketitle \normalsize \setcounter{footnote}{0} \vfill

@@

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
Examples of HLFs range from simple constructs such as counts or averages of
words, sentences or specific parts of speech~(adjectives, verbs, a.s.o) to more
complex statistics based on heuristics.
All HLFs share the characteristics of computational ease and idempotence.
This makes HLFs very attractive because of their dual potential.
On the one hand, an LLM might be instructed using a prompt containing the value
of an HLF.\@
An example prompt instruction which asks the LLM to conform to a maximum
\textit{sentence count} of 10 could be ``write no more than 10 sentences''.
On the other hand, HLFs can be computed on the text output by an LLM, too.
This dual nature of HLFs allows inspecting how closely does LLM generated text
conform to the goal outlined in the input prompt HLF instruction.\@

@@

We start by choosing experiment parameter values.
The following parameters have stable values throughout the experiment process: a
selection of HLFs, a corpus containing texts, a target LLM, examples from the
chosen corpus, one input text from the chosen corpus and one input text from
outside the chosen corpus.

Once the parameters have been determined, HLF statistics are computed on the
text corpus.
Remembering that HLFs are just floating point numbers, the following statistics
are computed for each HLF:~\textbf{min}~(the minimum value), \textbf{max}~(the
maximum value) and \textbf{avg}~(the mean).

For each HLF, an instruction is constructed in natural language, adhering to the
previously calculated minimum, maximum, and mean values. For instance, an
instruction corresponding to the total word count HLF might be expressed as
follows:

@@

The prompt templates correspond to each of the two experiment scenarios:
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The static instruction in `\texttt{prompt\_1}' to ``\textit{Write text that
conveys the meaning of the first user prompt}'' ensures that input text meaning
is preserved.
In addition to the primary directive from `\texttt{prompt\_1}',
`\texttt{prompt\_2}' contains instructions concerning the interpretation of the
provided examples, as well as logic to incorporate them.
Both prompt templates contain logic to distinguish between when HLF instructions
were given or not.

Baseline HLFs for the LLM output are computed after the preliminary steps are
completed.
For the first scenario, the LLM is instructed to reword the input text 10 times
by using `\texttt{prompt\_1}' without HLF instructions.
The HLFs in our selection are computed for each LLM output, resulting in a
10-dimensional vector for each HLF.\@
These vectors are the baseline HLFs for the LLM output in the first scenario.

The second scenario baseline HLFs are computed similarly to the first scenario
baselines, using the `\texttt{prompt\_2}' template and the examples provided as
an experiment parameter.
Remembering that the second scenario has two variants, we obtain two sets of
baseline vectors, one for each variant.
Note that the input from outside the corpus is the same as the input for the
first scenario, while the input from inside the corpus is selected along with
the corpus so that its HLFs are close to the average HLFs of the corpus.

From here on we refer to the HLFs computed using the above process as
\textit{baselines}.
We have 3 sets of baselines: 1 for the first scenario and 2 for the second
scenario.

For each set of baselines, we run 10 text generations using the corresponding
prompt augmented with the HLF instructions computed earlier in the process.
Doing this results in 3 sets of 10-dimensional vectors, each corresponding to
the baselines.
We refer to these vectors as \textit{HLF results}.

Lastly, we compare baselines with their corresponding HLF results for each HLF
in our selection, for each scenario and variant.

The comparison result for the first scenario indicates whether LLMs acknowledge
HLF instructions at all.
If they do, the HLF results should be significantly different from their
baseline.
Both variants of the second scenario quantify how much HLF instructions
influence text generation.
The underlying assumption is that LLMs are able to mimic examples when they are
provided in the system prompt.
In such a case, the presence of HLF instructions would remain among the few
possible explanations for deviations from the baseline, especially when the
baseline is already statistically close to the desired HLF values.

@@

The LLMs selected for the experiment must represent the state of the art.
The Chatbot Arena~\cite{chiang2024chatbot} leaderboard is essential in our
selection based on this criterion.
To cover more ground we encourage vendor diversity in our selection.
The final selection is available in Table~\ref{table-llm}.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
In this experiment we use LFTK~\cite{lftk-2023}, a framework built on top of
spaCy~\cite{spacy} which provides implementations for computing many HLFs.
LFTK categorizes HLFs by domain and family.
We select HLFs based on their domain and family from the catalog implemented by
LFTK, with the interest to make a diverse selection with respect to both
these attributes.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

@@

An input text from within each corpus is chosen so that it exhibits the closest
HLFs to the averages computed on the corpus.
The inputs are chosen from the corpora when the corpora are selected and are
stable throughout all experimental evaluations.

@@

HLFs are real numbers derived from text.
Plot figures show the HLF on the vertical axis and the text generation trial
number on the horizontal axis.
We plot the baseline using a continuous grey line and the HLF result using a
continuous grenat line.
The minimum and maximum HLF on the corpus are plotted using dotted black lines.
A continuous black line designates the average HLF on the corpus.
We refer to the corpus average as the \textit{target} because the LLM is
instructed to generate text that primarily conforms to this value.

@@

There are exceptions to the general trend observed for the first scenario.
One such exception is highlighted in Figure~\ref{fig-p2-gpt-twords}.
In~\cref{fig-p2-command-r-simpttr,fig-p2-gemini-a-kup-pw} we even observe
regressions in performance with significant differences between baseline and
result.
Even so, these are exceptions and the results obtained in our first scenario are
largely confirmed on the CNN article corpus.

@@

The results shown in Table~\ref{table-prompt-2-ifd-cnn-dailymail} contradict
these expectations.
The most surprising behaviour is exhibited for HLFs which are derived into more
concrete generation instructions, such as the total number of
words~(Figure~\ref{fig-p2-ifd-gemini-twords}).
Additionally, we notice regressive behaviour for some models compared to the
previously explored results in Figure~\ref{fig-p2-ifd-claude3-nuadj}.
Finally, we notice that the abstractness of the text generation instructions for
certain HLFs leads to non-compliance in Figure~\ref{fig-p2-ifd-llama3-a-kup-pw}.

@@

The overall sentiment, taking into account the additional results obtained on
the Yelp dataset, is that LLMs do not take much advantage of the examples
from text corpora, nor of the input from the corpora for the tasks performed in
this experiment.

@@

In terms of our first question --- whether HLF instructions impact an LLM's
writing, the response is affirmative.
Not all LLMs are equally susceptible to HLF instructions, though.
The experiment design involves choices that rely on the presence of a corpus and
the usage of a fixed external input text.
This is relevant as evidentiated by the difference in HLF results obtained using
the Yelp reviews corpus when compared to the HLF results obtained on the CNN
article corpus.
While similarly significant, the HLF results on Yelp reviews are mostly worse
than their corresponding baselines failing to confirm the positive impact of HLF
instructions.

The impact HLF instructions have on LLM output is limited.
Our metrics do not show a consistent desired impact on the LLM output either.
There isn't conclusive evidence that if we improve the LLM's chances of
producing a better baseline, this will result in closer HLF results to the 
desired target values.\@
In fact, there is some evidence that HLF instructions have an adverse effect
when examining the HLF results obtained on the Yelp reviews dataset.
This is especially true of highly abstract HLF instructions.
